A Framework for Reproducible Data Science using Nix

### 2025-10-17 paper.qmd
Update paper.qmd
getting ready

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order create an analysis that is reproducible, it is not enough to write
  clean code and document it well. One must also make sure to clearly list all
  the dependencies of the analysis and ideally provide an easy way to install
  them. There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study, such
  as [Docker]{.proglang}, a containerization solution. This paper will present
  another tool that can be used to make projects reproducible, the [Nix]{.pkg}
  package manager, and an [R]{.proglang} package called [rix]{.pkg} that makes
  using [Nix]{.pkg} much simpler for users of [R]{.proglang}.

@@

@peng2011 introduced the idea of reproducibility being on a continuum: on one of
the ends of this continuum, we only have access to the paper describing the
studies, which is not reproducible at all. If in addition to this paper authors
make the original source code of the analysis available, reproducibility is
improved. Adding the original data improves reproducibility yet again. Finally,
if to all this the authors add what Roger Peng referred to as the *linked and
executable code and data*, we reach the gold standard of reproducibility.

*Linked and executable code and data* refer to the compiled source code scripts,
but for the purposes of this article, we will assume this meas the
*computational environment*. The computational environment is all the software
required to run the analysis. Here too, we can speak of a continuum. One could
simply name and list the software used: for example, the [R]{.proglang}
programming language. Sometimes, authors have the courtesy to also state the
version of [R]{.proglang} used. Some authors go further, and also list the
packages used, and ideally with their versions as well. Authors rarely state the
operating system on which the analysis was done, even though it has been shown
that running the same analysis with the same software but on different operating
systems could lead to different results, as described in @neupane2019. Authors
also only very rarely provide instructions to install the required tools and
software in order to reproduce their studies.

But even assuming that authors provided clear instructions to install the exact
requirements of a study, it might happen that these instructions do not even
work on different platforms, or on different versions of the same platform. And
reproducibility is not the only activity that might get impacted by a lack of
clear instructions: if an author works on several machines, the exact same
computational environment should be used on all the machines, and the same is
true when a team of authors collaborate.

There are tools available that authors can use to reach the gold standard of a
fully reproducible study, or to successfully deploy the same development
environment on several machines. Let us first consider the task of listing the
software used. [R]{.proglang} provides the `sessionInfo()` function whose output
can be saved into a file. Below is an example output of `sessionInfo()`:

```{r, eval = FALSE}
sessionInfo()
```

```
R version 4.3.2 (2023-10-31)
Platform: aarch64-unknown-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

@@

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 [3] LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 [5] LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                 
 [9] LC_ADDRESS=C               LC_TELEPHONE=C            
[11] LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C

@@

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods
[7] base

other attached packages:
[1] nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163

loaded via a namespace (and not attached):
[1] compiler_4.3.2 Matrix_1.6-1.1 tools_4.3.2    splines_4.3.2
[5] grid_4.3.2     lattice_0.21-9
```

If an author provides this information, other people trying to reproduce the
study (or the author him- or herself in the future), or co-authors, can read
this file and see which version of [R]{.proglang} was used, and which packages
(and their versions) were used as well. However, others would still need to
install the correct versions of each package manually. This can be in certain
cases challenging, especially if packages that depend on system-level
dependencies are required. For example, the [nloptr]{.pkg} package requires a
binary of [nlopt]{.proglang} to be available, or if not, [cmake]{.proglang} on
Linux or macOS.

A better alternative to simply listing packages and their versions is to use the
[renv]{.pkg} package which generates a so-called `renv.lock` file which also
lists [R]{.proglang} and package versions. Here is an example of such an
`renv.lock` file:

```
{
"R": {
  "Version": "4.2.2",
  "Repositories": [
  {
   "Name": "CRAN",
   "URL": "https://packagemanager.rstudio.com/all/latest"
  }
  ]
},
"Packages": {
  "MASS": {
    "Package": "MASS",
    "Version": "7.3-58.1",
    "Source": "Repository",
    "Repository": "CRAN",
    "Hash": "762e1804143a332333c054759f89a706",
    "Requirements": []
  },
  "Matrix": {
    "Package": "Matrix",
    "Version": "1.5-1",
    "Source": "Repository",
    "Repository": "CRAN",
    "Hash": "539dc0c0c05636812f1080f473d2c177",
    "Requirements": [
      "lattice"
    ]

    ***lines below omitted***
```

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of [R]{.proglang}needs to be handled
separately. [renv]{.pkg} will also not install system-level dependencies such
as [cmake]{.pkg} for [nloptr]{.pkg} on Linux or macOS, so these still need to
be handled separately, as described in [renv]{.pkg}’s Readme: 
\url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

@@

```
groundhog.library("
    library(purrr)
    library(ggplot2)",
    "2017-10-04",
    tolerate.R.version = "4.2.2")
```

These packages also get installed in a project-specfic library so there is no
interference between these packages and other versions of the same packages that
one might use for other projects. Because [groundhog]{.pkg} does not install
[R]{.proglang} itself, users should either install the required version
themselves, or they should use the `tolerate.R.version` argument as shown in the
example above. Otherwise, [groundhog]{.pkg} would not continue with the
installation of the packages. Another alternative to [renv]{.pkg}, developed by
@chan2023 is [rang]{.pkg}, which also installs packages as they were on a given
date. Yet another way to install packages as they were on a give date is to use
the Posit Package Manager, which provides snapshots of CRAN. For example, to
install the required packages for an analysis as they were on the 30th of June
2023, one could add the following line to the `.Rprofile` file:

```
options(repos =
  c(REPO_NAME =
    "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
  )
)
```

The `.Rprofile` file gets read by [R]{.proglang}when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror. However, unless the user explicitly and
manually manages the different projects' libraries, using the Posit Package
Manager will install all the package for all the projects at the given date.

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.pkg} by the [R]{.proglang}infrastructure team [-@rlib2023]. [rig]{.pkg}
makes it easy to install different versions of [R]{.proglang}, so one could use
it to install the right version to reproduce an analysis, and then use one of
the listed packages above to install the right library of [R]{.proglang}
packages. However, this involves many manual steps and is thus error prone and
time-consuming.

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.pkg} image. [Docker]{.pkg} is a containerisation tool: using
[Docker]{.pkg} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that requires many software
dependencies. *Dockerizing* an analysis consists in first building an image: in
the build step, the dependencies of the analysis have to be installed and this
can be achieved using the tools mentioned above. The scripts to run the analysis
and the data are also added to the image at build time. These steps are written
down in a simple text file called a `Dockerfile`. This `Dockerfile` is the
definition of the [Docker]{.pkg} image.

It is possible to then execute the code of a study inside a so-called container,
which is a running instance of an image. Executing, or running, these containers
is done non-interactively, and so it is possible to instantiate the correct
environment to run a study in a single command.

The advantage of using [Docker]{.pkg} is not only about having the right version
of [R]{.proglang} and [R]{.proglang} packages bundled alongside the analysis
scripts: [Docker]{.pkg} images are actually almost complete Linux operating
systems, which means that system-level dependencies also get bundled into the
image. This is especially useful for packages that rely on external system-level
dependencies to run, such as for [nloptr]{.pkg}, as these external system-level
dependencies are thus also made available to future replicators of the study
within the [Docker]{.pkg} image. The image can also be shared.

The Rocker project initiated by @boettiger2017 provides many pre-built
[Docker]{.pkg} to the [R]{.proglang} community of users, already containing
specific [R]{.proglang} versions and even packages. These images can be used as
bases to build other images containing statistical analyses more easily than
starting from a bare-bones image.

While [Docker]{.pkgs} offers a great solution to the reproducibility problem, it
can be awkward to work interactively from a running container: containers can be
changed at run-time, but any change will be lost once the container is stopped.
Also, while it is possible to launch graphical programs from a container, this
is generally not easy to set up and only works, as far as we know, only on Linux
operating systems. In most cases, users of [Docker]{.pkg} will instead use the
container to run a web server and the web based application of a popular
integrated development environment (IDE).

A common approach seems to be to work on a study interactively using a standard
installation of [R]{.proglang}and [R]{.proglang}packages, and after the study is done, take some time to
write the required `Dockerfile` to provide an easy way for future replicators to
reproduce the results of the study. But this approach does not solve the problem
of deploying the same environment on several machines, and so ideally authors
would develop using a containerized development environment. Another challenge
of using [Docker]{.pkg} is that, as stated already, [Docker]{.pkg} images are
actually almost complete Linux operating systems, so experience with Linux is
highly recommended if one wishes to properly write a `Dockerfile`. Finally, one
also needs to be careful and write the `Dockerfile` in such a way that the build
process will always result in the same image. This is only possible if a
specific digest of a base image is used instead of a tagged version, but in
practice, this is rarely done.

A final, and optional, step towards the gold standard is to use a build
automation tool such as [Make]{.pkg} to run the whole analysis when the
container is executed, instead of executing scripts. Build automation tools make
it easier to run arbitrary code in a series of well-defined steps.
[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
automation tool.

@mcdermott2021 is an example of a scientific study that reached the gold
standard of reproducibility. The author of this paper set up an accompanying
Github repository to the
paper^[https://github.com/grantmcdermott/skeptic-priors] containing all the
instructions to install the required software and then run the analysis. If we
take a closer look at this repository, we will notice that many of the tools
previously mentioned were used to capture the compatutational environment and
make it available to other researchers:

- Packages and their versions were listed and saved into an `renv.lock` file;
- A `Makefile` was used to run the whole analysis and compile the paper;
- A `Dockerfile` was used to provide the complete computational environment,
  including the right version of [R]{.proglang}, to run the whole analysis
  easily.

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.pkg} for all the other software,
including the programming language that was used. Ideally, a build automation
tool such as [Make]{.pkg} should be added to the list. It should also be noted
that these requirements are not limited to the [R]{.proglang} programming
language. A very similar approach should be taken if one uses
[Python]{.proglang}, or any other programming language. Furthermore, the
previous discussion also implies that for an analysis to actually be
reproducible at all, it must be made using free and open source software, as
proprietary software cannot be freely shared with others that wish to reproduce
a study, and there's no guarantee that others will have access to the
proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.pkg} package
manager, which is available for all major operating systems and focuses on
installing and building software in a reproducible manner. This means that this
single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg} and even a
build automation tool such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
to [R]{.proglang} programmers, we developed the [rix]{.pkg} package which we
will present in this article.

@@

[Nix]{.pkg} is a package manager that can be used to install and build software
in a completely reproducible manner. As of writing, it contains more than
100.000 packages, including almost the entirety of CRAN and Bioconductor. This
means that using [Nix]{.pkg}, it is possible to install not only [R]{.proglang},
but also all the packages required for a project. [Nix]{.pkg} is actually the
package manager of a Linux distribution called NixOS, but can be installed as a
stand-alone tool for other Linux distributions, macOS, and on the Windows
Subsystem for Linux on Windows^[Which means that in practical terms, non-Nixos
Linux distributions and Windows can be considered as the same platform.].

The reason why one should use [Nix]{.pkg} to install [R]{.proglang} packages,
and not use the usual, built-in, `install.packages()` [R]{.proglang} function
instead, is that when installing a package with [Nix]{.pkg}, [Nix]{.pkg} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not be easy, nor even possible. But with [Nix]{.pkg}, it
suffices to declare that the [xlsx]{.pkg} package is needed for the project, and
[Nix]{.pkg} will install [Java]{.proglang} and configure it. This is possible
because the maintainers of the [R]{.proglang} language for [Nix]{.pkg} declared
[Java]{.proglang} as a dependency of [xlsx]{.pkg}. From the perspective of the
end-users, this all happens automatically, thanks to the efforts of the
maintainers.

@@

But how does [Nix]{.pkg} achieve this? Where do these packages, or closures,
come from? When installing a package using [Nix]{.pkg}, an expression written in
the [Nix]{.pkg} programming language gets downloaded from the `nixpkgs` Github
repository and evaluated. `nixpkgs` is the name of the repository that contains
the more than 100'000 pieces of software that can be installed with the
[Nix]{.pkg} manager. Expressions contain so-called *derivations*. A derivation
defines a build: its dependencies, commands to build and install the package in
question, and then an output. Most of the time, a derivation downloads source
code, builds the software from the source and then outputs a compiled binary.
Derivations are extremely flexible, and by writing one's own, it is possible to
define and build an environment for a project in a reproducible manner. The
dependencies of these derivations are also defined in other expressions, which
will also get installed if needed. As explained above, the maintainers of
[R]{.proglang} for [Nix]{.pkg} declare the right dependencies for the packages
to turn them into component closures.

But why is installing software with [Nix]{.pkg} reproducible? Because the whole
set of [Nix]{.pkg} expressions is hosted on GitHhub, it is possible to use a
specific commit of `nixpkgs` to ensure reproducibility of our project. Using a
specific commit of `nixpkgs` is called *pinning a revision (or commit)*. Pinning
a revision ensures that every package that [Nix]{.pkg} installs will always be at
exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.pkgs} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`.

While pinning revisions is essential for reproducibility, this is not the only
reason [Nix]{.pkg} is very useful for reproducibility, but the [Nix]{.pkg}
package manager itself was designed with reproducibility in mind. [Nix]{.pkg} is
a functional package manager, meaning that it uses concepts from functional
programming to build software: no-side effects or global variables are allowed,
and the output is always the same if given the same inputs, regardless of where
or when a package gets built (an important side-note is in order here: while the
functional aspect of [Nix]{.pkg} is quite useful to ensure reproducibility, it
can lead to situations where writing the required [Nix]{.pkg} expression to
build it can be much more difficult than usual: this is especially true for
packages that must download assets from the internet during installation, such
as several Bioconductor packages like [musData]{.pkg} for example. But this is
only a concern for the packagers, not users of the package that simply wish to
install and use the package). [Nix]{.pkg} also allows for multiple versions (or
variants in the [Nix]{.pkg} jargon) to be installed at the same time on the same
system. Each variant of a package has a unique identifier, so that it is
possible to have two versions of [R]{.proglang} installed and use the correct
one where adequate. We refer to @dolstra2004nix for a more technical discussion
on [Nix]{.pkg}.

>>>>

---
title: "A Framework for Reproducible Data Science using Nix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
abstract: |
  To create a reproducible analysis, it is not enough to write clean,
  well-documented and tested code. One must also ensure that all software
  dependencies are clearly declared and, ideally, provide a simple way to
  install them. Several tools exist for this purpose, such as the
  containerization solution [Docker]{.pkg}. A complete solution, however,
  requires not only managing the computational environment but also
  orchestrating the execution of the analysis itself in a robust and automated
  way. This paper presents a comprehensive framework based on the [Nix]{.pkg}
  package manager to solve both problems. We first introduce the [R]{.proglang}
  package [rix]{.pkg}, which simplifies the creation of reproducible
  environments with [Nix]{.pkg}. We then introduce [rixpress]{.pkg}, a second
  [R]{.proglang} package that leverages these environments to define and execute
  complex, multi-language analytical pipelines. Together, these tools provide a
  powerful, unified approach to achieving high standards of computational
  reproducibility.

@@

@peng2011 introduced the concept of reproducibility as a *continuum*. At one end
lies the least reproducible state, where only a paper describing the study is
available. Reproducibility improves when authors share the original source code,
improves further when they include the underlying data, and reaches its highest
level when what Roger Peng called *linked and executable code and data* are
provided.

By *linked and executable code and data*, Peng referred to compiled source code
and runnable scripts. In this paper, we interpret this notion more broadly as
the *computational environment*: the complete set of software required to
execute an analysis. Here too, a continuum exists. At the minimal end, authors
might only name the main software used—say, the [R]{.proglang} programming
language. More careful authors might also specify the version of [R]{.proglang},
or list the additional packages and their versions. Rarely, however, do authors
specify the operating system on which the analysis was performed, even though
differences in operating systems can lead to divergent results when using the
same code and software versions, as shown by @neupane2019. It is even less
common for authors to provide step-by-step installation instructions for the
required software stack.

Even when such instructions are given, they often fail across different
platforms or versions of the same platform. This lack of portability not only
hinders reproducibility but also complicates everyday research workflows.
Researchers working across multiple machines must be able to recreate the same
environment consistently, and collaborators must share identical computational
setups to avoid inconsistencies.

Finally, once the execution environment is correctly configured, additional
clarity is needed on how to *run* the project itself. Which packages should be
loaded first? Which scripts should be executed, and in what order? Without clear
documentation or a logically organized project structure, these operational
details become yet another barrier to reproducibility.

A range of tools now exist to help researchers approach the gold standard of
full reproducibility, or to consistently deploy the same development environment
across multiple machines. Let us first consider the most basic step in this
process: listing the software used. In [R]{.proglang}, the `sessionInfo()`
function provides a concise summary of the software environment, including the R
version, platform details, and all loaded packages. Its output can be saved to a
file and included as part of a study’s reproducibility record. Below is an
example output from `sessionInfo()`:

```{=tex}
\begin{CodeInput}
R> sessionInfo()
\end{CodeInput}
\begin{CodeOutput}
R version 4.3.2 (2023-10-31)
Platform: aarch64-unknown-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

@@

locale:
 LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C              
 LC_TIME=en_US.UTF-8        LC_COLLATE=en_US.UTF-8    
 LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 LC_PAPER=en_US.UTF-8       LC_NAME=C                 
 LC_ADDRESS=C               LC_TELEPHONE=C            
 LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C

@@

attached base packages:
 stats     graphics  grDevices utils     datasets  methods   base

other attached packages:
 nnet_7.3-19  mgcv_1.9-0   nlme_3.1-163

loaded via a namespace (and not attached):
 compiler_4.3.2  Matrix_1.6-1.1  tools_4.3.2     splines_4.3.2
 grid_4.3.2      lattice_0.21-9 
\end{CodeOutput}
```

When an author includes this information, others attempting to reproduce the
study (future readers, collaborators, or even the author at a later
time) can easily see which version of [R]{.proglang} and which packages (with
their versions) were used. However, reproducing the environment still requires
manually installing the correct package versions. This can be difficult,
especially when packages depend on system-level libraries. For example, the
[nloptr]{.pkg} package requires a precompiled [nlopt]{.proglang} binary, or
alternatively, [cmake]{.proglang} when building from source on Linux or macOS.

A more robust approach than simply listing package versions is to use the
[renv]{.pkg} package. [renv]{.pkg} captures the project’s software state and
writes it to a `renv.lock` file, which includes the exact versions of
[R]{.proglang} and all required packages. This lockfile serves as a blueprint
for restoring the environment automatically, ensuring that others can recreate
the same setup with minimal effort. Below is an example of an `renv.lock` file:

```{=tex}
\begin{Code}
{
  "R": {
    "Version": "4.2.2",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://packagemanager.rstudio.com/all/latest"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-58.1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "762e1804143a332333c054759f89a706",
      "Requirements": []
    },
    "Matrix": {
      "Package": "Matrix",
      "Version": "1.5-1",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "539dc0c0c05636812f1080f473d2c177",
      "Requirements": [
        "lattice"
      ]
    }

    ... lines below omitted ...
  }
}
\end{Code}
```

This lockfile lists each package alongside its version and the repository from
which it was downloaded. Creating it requires only a single command:
`renv::init()`. Others can then reproduce the same package library by running
`renv::restore()`, which installs the exact package versions in an isolated,
project-specific library that does not interfere with the user’s global R
library.

However, [renv]{.pkg} does not restore the version of [R]{.proglang} itself.
Installing the correct R version must therefore be done separately. Similarly,
`renv` does not handle system-level dependencies such as [cmake]{.proglang},
which is required by [nloptr]{.pkg} on Linux and macOS. These dependencies must
be managed manually or through other tools. Further details are discussed in the
[renv]{.pkg} documentation: \url{https://rstudio.github.io/renv/articles/renv.html#caveats}.

@@

```{=tex}
\begin{CodeInput}
R> groundhog.library("
+   library('purrr')
+   library('ggplot2')",
+   "2017-10-04",
+   tolerate.R.version = "4.2.2")
\end{CodeInput}
```

Packages installed with [groundhog]{.pkg} are placed in a project-specific
library, ensuring they do not interfere with other versions of the same packages
used in different projects. Since [groundhog]{.pkg} does not install
[R]{.proglang} itself, users must either install the required R version manually
or use the `tolerate.R.version` argument, as illustrated in the example above.
Without this argument, [groundhog]{.pkg} will not proceed with package
installation if the R version does not match the expected one.

An alternative to [renv]{.pkg} is [rang]{.pkg}, developed by @chan2023, which
similarly allows users to install packages as they existed on a specific date.
Another approach is to use the Posit Package Manager, which provides dated
snapshots of CRAN. For instance, to ensure that packages are installed as they
were on June 30, 2023, one could include the following line in the `.Rprofile`
file:

```{=tex}
\begin{CodeInput}
R> options(repos =
+    c(REPO_NAME =
+      "https://packagemanager.posit.co/cran/__linux__/jammy/2023-06-30"
+    )
+  )
\end{CodeInput}
```

The `.Rprofile` file is read by [R]{.proglang} at the start of each new session,
meaning that every call to `install.packages()` will install packages from the
specified snapshot mirror. However, unless users explicitly manage separate
libraries for different projects, the Posit Package Manager will install all
packages for all projects as they existed on that snapshot date.

The next step toward the gold standard of reproducibility is ensuring not only
that the correct package versions are installed, but also that the correct
version of [R]{.proglang} itself is used. While this can be done manually,
dedicated tools make the process more efficient. One such tool is [rig]{.pkg},
developed by the [R]{.proglang} Infrastructure Team [-@rlib2023]. [rig]{.pkg}
simplifies installing and managing multiple R versions, allowing users to match
the exact version needed for a given analysis. Once the correct R version is
installed, one can then use tools such as [renv]{.pkg}, [groundhog]{.pkg}, or
[rang]{.pkg} to recreate the associated package library.

However, this multi-step process remains manual, error-prone, and
time-consuming, highlighting the need for more comprehensive and automated
solutions to achieve full computational reproducibility.

The final tool for achieving the gold standard of reproducibility is to bundle
the exact version of [R]{.proglang} and its packages within a [Docker]{.pkg}
image. [Docker]{.pkg} is a containerization platform that allows packaging a
*data product* together with all its dependencies into a self-contained image. A
statistical analysis can be viewed as such a data product, requiring a specific
set of software dependencies to run. *Dockerizing* an analysis involves building
an image that installs these dependencies, often using the tools mentioned
earlier, and includes the analysis scripts and data (but data can also be
dynamically made available at run-time, to avoid having to copy it inside the
image). The steps to create this image are specified in a simple text file
called a `Dockerfile`, which defines the structure and content of the
[Docker]{.pkg} image.

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. Containers are typically run
non-interactively, allowing the complete computational environment to be
instantiated and executed with a single command.

The strength of [Docker]{.pkg} lies not only in its ability to encapsulate the
correct versions of [R]{.proglang} and its packages but also in its inclusion of
system-level dependencies. Since [Docker]{.pkg} images are nearly complete Linux
systems, they naturally bundle libraries and tools required by certain R
packages (for example, [nloptr]{.pkg}’s dependency on the [nlopt]{.proglang}
library). This ensures that future replicators of a study have access to the
same software environment, including all necessary system components. Moreover,
these images can be easily shared, enabling seamless reproducibility across
machines and collaborators.

The Rocker project, introduced by @boettiger2017, provides a collection of
pre-built [Docker]{.pkg} images for the [R]{.proglang} community. These images
come with specific R versions and, in some cases, preinstalled packages, making
them convenient base images for building reproducible analysis environments
without starting from scratch.

Despite its strengths, [Docker]{.pkg} can be cumbersome for interactive use.
Although containers can be modified at runtime, any changes are lost once the
container stops unless they are explicitly saved. Running graphical applications
from containers is also possible, but it is generally difficult to configure and
tends to work reliably only on Linux systems. For this reason, [Docker]{.pkg} is
most often used to run web-based applications (such as integrated development
environments, IDEs, like the web version of RStudio) inside containers, offering
a reproducible yet flexible environment for interactive work.

Some IDEs provide built-in support for executing code directly within a running
container, allowing a more seamless workflow. However, these setups typically
require additional configuration, and many general-purpose IDEs are not well
optimized for statistical or data science workflows (though Positron may soon
change that).

A common practice is to conduct a study interactively using a standard
installation of [R]{.proglang} and its packages, and only after completing the
analysis, create a `Dockerfile` to allow others to reproduce the results easily.
For instance, the [dockerfiler]{.pkg} package [@dockerfiler] can automatically
generate a `Dockerfile` from an existing `renv.lock` file, simplifying this
process. While this approach enables post-hoc reproducibility, it does not solve
the challenge of consistently deploying the same environment across multiple
machines as co-authors are working on the project. Ideally, analyses would be
developed directly within a containerized environment from the start, ensuring
that the computational setup is reproducible throughout the project lifecycle.

Another challenge of using [Docker]{.pkg} is that its images are effectively
minimal Linux systems, so familiarity with Linux is highly recommended for
writing reliable and efficient `Dockerfile`s. To guarantee that a build
consistently produces the same image, the `Dockerfile` should reference specific
image *digests* rather than general *tags*. In practice, however, tagged
versions are more commonly used, which can introduce potential reproducibility
issues. [dockerfiler]{.pkg} is an R package that lowers the
barrier to entry by automating parts of the `Dockerfile` creation process, but
they cannot fully replace the need for basic Linux knowledge.

A final step toward the gold standard of reproducibility is to automate the
analysis execution using a build system such as [Make]{.pkg}, rather than
relying on ad hoc script execution. Build automation tools enable analyses to be
defined as a series of reproducible, well-ordered steps that can be executed
consistently. Within [R]{.proglang}, a modern and powerful alternative is the
[targets]{.pkg} package by @landau2021, which provides declarative workflows for
reproducible data analysis.

@mcdermott2021 provides a clear example of a scientific study that achieved the
gold standard of reproducibility. The author created an accompanying GitHub
repository^[https://github.com/grantmcdermott/skeptic-priors] containing
detailed instructions to install the required software and execute the analysis.
Examining this repository reveals the use of multiple tools to capture and share
the computational environment:

- Packages and their versions were recorded in an `renv.lock` file;
- A `Makefile` was used to automate the full analysis and compile the paper;
- A `Dockerfile` provided a complete computational environment, including the
  correct version of [R]{.proglang}, enabling easy execution of the entire
  workflow.

Reaching this gold standard, however, is resource-intensive. It requires
learning a tool for managing package versions, mastering [Docker]{.pkg} for the
programming language and other software dependencies, and using a build
automation tool. Importantly, these considerations are not unique to
[R]{.proglang}: similar approaches are needed for [Python]{.proglang} or any
other programming language, and complexity further increases if several
programming languages are needed for a single project.

As an alternative, we present the [Nix]{.pkg} package manager, which supports
all major operating systems and emphasizes reproducible software installation
and building. [Nix]{.pkg} can replace [Docker]{.pkg}, [renv]{.pkg}, and even
build automation tools such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
for [R]{.proglang} users, we developed the [rix]{.pkg} and [rixpress]{.pkgs}
packages, which are introduced and discussed in this article.

@@

[Nix]{.pkg} is a package manager designed to install and build software in a
fully reproducible manner. As of this writing, it provides access to over
120000 packages, including nearly all of CRAN and Bioconductor. This allows
users to install not only [R]{.proglang} itself but also all packages required
for a given project. Although [Nix]{.pkg} is the package manager of the NixOS
Linux distribution, it can also be installed as a standalone tool on other Linux
distributions, macOS, and the Windows Subsystem for Linux^[Effectively treating
non-NixOS Linux distributions and Windows as equivalent platforms in practice.].

The advantage of using [Nix]{.pkg} to install [R]{.proglang} packages, instead
of the standard `install.packages()` function, is that [Nix]{.pkg} ensures all
dependencies of each package are installed, whether they are other
[R]{.proglang} packages or system-level libraries.

For example, the [xlsx]{.pkg} package requires [Java]{.proglang} to be
installed. On some systems, installing [Java]{.proglang} may be difficult or
even impossible manually. With [Nix]{.pkg}, however, the user only needs to
declare that [xlsx]{.pkg} is required for the project; [Nix]{.pkg} automatically
installs and configures [Java]{.proglang} as a dependency. This works because
the maintainers of the [R]{.proglang} language packages for [Nix]{.pkg} have
declared [Java]{.proglang} as a dependency of [xlsx]{.pkg}, allowing the process
to happen seamlessly from the end-user perspective.

@@

But how does [Nix]{.pkg} achieve this reproducibility? Where do these packages,
or *closures*, come from? When installing a package with [Nix]{.pkg}, an
expression written in the [Nix]{.pkg} language is downloaded from the `nixpkgs`
GitHub repository and evaluated. These expressions define *derivations*, which
describe how to build a package: specifying its dependencies, the commands to
build and install it, and the resulting output. Typically, a derivation
downloads the source code, compiles it, and outputs a binary. Derivations are
highly flexible; by writing custom derivations, users can define and build
reproducible environments instead of a single program. The dependencies of each
derivation are themselves defined in other expressions and are automatically
installed if needed. In the case of [R]{.proglang}, maintainers declare the
appropriate dependencies to create fully self-contained component closures.

Why is installing software with [Nix]{.pkg} reproducible? Because the entire set
of [Nix]{.pkg} expressions is hosted on GitHub, it is possible to reference a
specific commit of `nixpkgs`, a process called *pinning a revision*. Pinning a
revision ensures that all packages installed by [Nix]{.pkg} are always the exact
same versions, regardless of when in the future they are built, since the
expressions downloaded correspond to that specific commit of `nixpkgs`.

Pinning is crucial, but it is not the only reason [Nix]{.pkg} supports
reproducibility. [Nix]{.pkg} is a functional package manager: it applies
principles from functional programming, disallowing side effects and global
variables, and ensuring that the output of a build is always the same given the
same inputs, regardless of time or location. A side note: while this functional
approach greatly enhances reproducibility, it can make writing derivations more
complex, particularly for packages that need to download assets during
installation (e.g., some Bioconductor packages like [musData]{.pkg}). However,
this concern is primarily for package maintainers, not end-users.

Additionally, [Nix]{.pkg} supports multiple versions, or *variants*, of a
package on the same system. Each variant has a unique identifier, allowing
multiple versions of [R]{.proglang} to coexist and ensuring the correct version
is used for each project. For a more technical discussion of [Nix]{.pkg}, we
refer to @dolstra2004nix.

With [Nix]{.pkg}, it is possible to effectively replace both [renv]{.pkg} and
[Docker]{.pkg} for [R]{.proglang} projects—or, in the case of
[Python]{.proglang}, to replace `requirements.txt` files and virtual
environments. [Nix]{.pkg} also supports building multi-language environments,
which can include [R]{.proglang} and [Python]{.proglang}, a \LaTeX
distribution or any other of the 120,000 available tools. This enables the
creation of a complete, project-specific, and reproducible environment that can
be used interactively or non-interactively. As long as the `nixpkgs` GitHub
repository remains online, the environment can be rebuilt in the future to rerun
analyses reliably.

However, [Nix]{.pkg} has a steep learning curve. It is a complex system that
comes with its own programming language, also called [Nix]{.proglang}, designed
to solve the challenging problem of declaratively defining how software is built
and configured. This ensures that installations are fully reproducible across
operating systems and hardware platforms. To make [Nix]{.pkg} more accessible to
[R]{.proglang} users that wish to use project-specific and reproducible
development environments, we developed the [rix]{.pkg} package.

## Reproducible development environments with Nix {#sec-repro-nix}

As mentioned, [Nix]{.pkg} expressions are written in the [Nix]{.proglang}
programming language, which is purely functional. Here is a simple example that
creates a shell environment containing version 4.3.1 of [R]{.proglang}:

```{=tex}
\begin{CodeInput}
let
  pkgs = import (fetchTarball
    "https://github.com/NixOS/nixpkgs/archive/976fa336.tar.gz"
  ) {};
  system_packages = builtins.attrValues {
    inherit (pkgs) R;
  };
in
  pkgs.mkShell {
    buildInputs = [ system_packages ];
    shellHook = "R --vanilla";
  }
\end{CodeInput}
```
